Allee Effect and Cultural Evolution

### 2015-05-28 manuscript/draft.tex

Several cultural traits are expected to show such density dependence. In particular, behaviour linked to the exploitation of limited resources are expected to exhibit a negative density dependence. Novel optimal variants are likely to spread rapidly when the trait is rare, but once a threshold frequency is exceeded (i.e. the carrying capacity),  beneficial effect will start to decline, the payoff will become smaller, and eventually rarer suboptimal variants can be preferred (see Lake and Crema 2012 for an extensive analysis).

However, density dependence does not necessarily imply only a negative correlation between population size and payoff/fitness. In many ecological contexts an inverse density dependence, whereby per capita growth rate is positively correlated with population density, has been noted. This phenomena, known as Allee effect \citep{allee1958,courchamp_etal_1999}, is generally explained by a range of processes from cooperation to unconscious mutual facilitation that enhances individual fitness at low population density. This inverse density dependence can be further distinguished between \emph{weak} and \emph{strong} Allee effect. In the former case, the reproductive fitness is always positive at low population density, whilst in the latter case we expect the presence of a critical density below which fitness becomes negative \citep{Wang_and_Kot_2001}.

Variants that are enhanced by some form of cooperation, or more in general traits exhibiting a positive niche construction are expected to show such a pattern, known in ecology as \emph{Allee Effect}~\citep{allee1958}.

While a systematic review of the empirical evidence of Allee effect has been conducted in population ecology \citep{kramer_etal_2009}, no

For example foraging groups are known to show higher performance when the number of individuals is not too small~\citep{hill_and_hawkes_1983,janssen_and_hill_2014}, while certain activities such as anthropogenic fire~\citep{bird2013} or selective hunting~\citep{dods_2002} might require a population density beyond a certain threshold in order to promote sufficient niche enhancement and a consequent increase in payoff~\citep{rowley-conwy_and_layton_2011}. Other traits that might be promoted by increased density includes language~\citep{kandler2009}, organisational populations~\citep{caroll_and_hannan_1989} and communication technologies~\citep{van_slyke_perceived_2007}.

%We definitely need to add some literature review on the theorethical models of Allee effect here as an additional paragraph.

Here we explore the combined effect of positive and negative density-dependent selection, assuming that: 1) the social learning strategy is payoff-based transmission; and 2)  payoff is equivalent to the reproductive fitness of each individual. We will demonstrate, by exploring a two-trait difference equation, that the impact of Allee effect on social learning can strongly affect the evolutionary trajectory of the system leading to the emergence of stable and unstable equilibria, as well as the occurrence of temporary or permanent episodes of reversion to suboptimal variants.

\section{The Model}
We consider the evolutionary dynamics of two mutually exclusive subsistence strategies whose fitness has both positive and negative density dependence. In particular we examine how two population of individuals, $a$ and $b$, change over time if we allow for a social learning strategy guided by a \emph{proportional imitation rule} (i.e. the probability of an individual with lower payoff to adopt the trait of an individual with higher payoff is proportional to the difference in the payoff). The model can theoretically be extended to $n$ strategies, but here we decide to explore a two trait version for simplicity. The core model can be depicted with the coupled difference equation \eqref{eq1}:

@@

where $a_{t+1}$ and $b_{t+1}$ are the population using each trait at time $t+1$. This is given by the number of individuals during the previous time-step ($a_t$ and $b_t$), the reproductive rate of each population ($R_{a,t}$ and $R_{b,t}$), and the shift from one population to another dictated by cultural transmission ($C_{a,b,t}$).
% XAVI do you want to use cultural trasmission here or better stick with social learning?
The reproductive growth rate $R$ of each population is defined by the following paired equation:

@@

where $r$ is the basic growth rate, $A$ is the \emph{Allee} threshold and $K$ is the carrying capacity, following the condition $A < K$. Equation \eqref{eq2} implies that the reproductive rate is negative when density is below $A$ or above $K$, and will reach its maxima $R_{max}=\frac{r(A-K)2}{4AK}$ when the population density is equal to $\frac{(K-A)}{2}$.

Cultural transmission is portrayed in \eqref{eq3}:

@@

% XAVI b is not explained?
where $z$ is a measure of reliance in social learning (effectively equivalent to highest possible transmission rate), $\Delta$ is the maximum between $R_{max(a,t)}$ and $R_{max(b,t)}$, and $m$ is a calibration parameter that measure the perception of the difference in the payoffs (i.e. small values of $m$ determines a higher rate by which the theoretical maximum transmission rate is reached). Thus \eqref{eq3} and \eqref{eq4}  conforms with the \emph{proportional imitation rule}~\citep{schlag1998}, although here the payoff is equivalent to the reproduction rate.

@@

>>>>

Several cultural traits are expected to show such density dependence. In particular, behaviour linked to the exploitation of limited resources are expected to exhibit a inverse density dependence. Novel optimal variants are likely to spread rapidly when the trait is rare, but once a threshold frequency is exceeded (i.e. the carrying capacity),  beneficial effect will start to decline, the payoff will become smaller, and eventually rarer suboptimal variants can be preferred (see Lake and Crema \citeyear{lake_and_crema_2012} for an extensive analysis).

However, density dependence does not necessarily imply only a negative correlation between population size and payoff/fitness. In many ecological contexts an inverse density dependence, whereby per capita growth rate is positively correlated with population density, has been noted. This phenomena, known as Allee effect \citep{allee1958,courchamp_etal_1999}, is generally explained by a range of processes from cooperation to unconscious mutual facilitation that enhances individual fitness at low population density. This inverse density dependence can be further distinguished between \emph{weak} and \emph{strong} Allee effect. In the former case, the reproductive fitness is always positive at low population density, whilst in the latter case we expect the presence of a critical density below which fitness becomes negative \citep{Wang_and_Kot_2001}. The implications of direct and inverse density dependence has been extensively studied both from the empirical and theoretical perspectives, including its effect of habitat selection \citep{greene_and_stamps_2001}, resource competition \citep{jang2013}, and dispersal \citep{steele_2009}, as well as specific events such as shifts in settlement pattern \citep{crema_2014}, colonisation events,  \citep{kennet_etal_2006}, genetic consequences on expanding population \citep{roques_etal_2012} and extinction of languages \citep{sutherland_2003}

Although a systematic review and a shared theoretical framework akin to those proposed in population ecology \citep{kramer_etal_2009} is not present in the social sciences, these studies suggest that there are several lines of evidence supporting the existence of both direct and inverse density dependence in cultural and behavioural traits. For example several ethnographic studies on small scale societies show how the per-capita return of hunting groups show higher yield when the group size is not too small nor too big~\citep{hill_and_hawkes_1983,janssen_and_hill_2014}. Other instances of inverse density dependence include population level activities such as the use of anthropogenic fire~\citep{bird2013} or selective hunting~\citep{dods_2002} which often require a population size above a certain critical density in order to promote sufficient niche enhancement and a consequent increase in payoff~\citep{rowley-conwy_and_layton_2011}. In a broader sense any cultural trait that benefits from direct or indirect cooperation \citep{}, promote positive niche construction \citep{vandermeer_2008}, and requires a critical mass \citep{rogers_2003} are likely to exhibit some degree of Allee effect. This is thus not limited necessarily to the domain of traits directly promoting reproductive fitness, and hence can include language~\citep{kandler2009}, organisational populations~\citep{caroll_and_hannan_1989} and communication technologies \citep{van_slyke_perceived_2007}.

Here we explore the consequences of Allee effect in the evolutionary dynamics of cultural transmission, assuming that: 1) the social learning strategy is payoff-based; and 2)  the payoff is equivalent to the reproductive fitness of each individual. We will demonstrate, by means of a difference equation, that direct and inverse density dependence can strongly affect the evolutionary trajectory of the system leading to the emergence of stable and unstable equilibria, as well as the occurrence of temporary or permanent episodes of reversion to suboptimal cultural variants.

\section{The Model}
We consider the evolutionary dynamics of two mutually exclusive cultural variants whose payoff function is characterised by both direct and inverse density dependence. We assume a non-fixed population of size $N$, subdivided into two subpopulations with sizes $a$ and $b$, each associated with the two cultural variants. We further assume that $a$ and $b$ can change as a result of intrinsic population growth/decline as well as shifts from one to the other as a result of cultural transmission based. The core model can be depicted with the following coupled difference equation \eqref{eq1}:

@@

where $a_{t+1}$ and $b_{t+1}$ are the population using each trait at time $t+1$. This is given by the number of individuals during the previous time-step ($a_t$ and $b_t$), the reproductive rate of each population ($R_{a,t}$ and $R_{b,t}$), and the shift from one population to another dictated by cultural transmission ($C_{a,b,t}$).
% XAVI do you want to use cultural trasmission here or better stick with social learning?

The reproductive growth rate $R$ of each population is defined by the following paired equation:

@@

where $r$ is the basic growth rate, $A$ is the critical density and $K$ is the carrying capacity, following the condition $0<A < K$. Equation \eqref{eq2} depicts a \emph{strong} Allee effect, hence the reproductive rate is negative when density is below $A$ or above $K$, and will reach its maxima $R_{max}=\frac{r(A-K)2}{4AK}$ when the population density is equal to $\frac{(K-A)}{2}$.

Cultural transmission is based on the \emph{proportional imitation rule} \citep{schlag1998}, and hence in this case the probability of an individual with lower growth rate to adopt the trait of an individual with higher growth rate is proportional to the difference in the growth rates. This is portrayed in \eqref{eq3}:

@@

% XAVI b is not explained? %It's no longer b but m
where $z$ is a measure of reliance in social learning (effectively equivalent to highest possible transmission rate), $\Delta$ is the maximum between $R_{max(a,t)}$ and $R_{max(b,t)}$, and $m$ is a calibration parameter that measure the perception of the difference in the payoffs (i.e. small values of $m$ determines a higher rate by which the theoretical maximum transmission rate is reached).

@@

The results
